# Matrix algebra, very briefly

Matrix algebra is well-suited for ecology, because most (if not all) _data sets_ we work with are in a _matrix_ format.

## Data sets _are_ matrices

Ecological data tables are obtained as object-observations or sampling units, and are often recorded as this: 

| Objects | $y_1$     | $y_2$  | $\dots$           | $y_n$  |
| :-------------: |:-------------:| :-----:|:-----:|:-----:|
|  $x_1$        | $y_{1,1}$     | $y_{1,2}$  | $\dots$  | $y_{1,n}$  |
|  $x_2$        | $y_{2,1}$     | $y_{2,2}$  | $\dots$  | $y_{2,n}$  |
|  $\vdots$     | $\vdots$     | $\vdots$  | $\ddots$  | $\vdots$  |
|  $x_m$        | $y_{m,1}$     | $y_{m,2}$  | $\dots$  | $y_{m,n}$  |

where $x_m$ is the sampling unit $m$; and $y_n$ is the ecological descripor that can be, for example, species present in a sampling unit, locality, or a chemical variable.

The same ecological data table can be represented in _matrix notation_ like this:

$$Y = [y_{m,n}] =
\begin{bmatrix}
y_{1,1} & y_{1,2} & \cdots & y_{1,n} \\
y_{2,1} & y_{2,2} & \cdots & y_{2,n} \\
\vdots  & \vdots  & \ddots & \vdots  \\
y_{m,1} & y_{m,2} & \cdots & y_{m,n} 
\end{bmatrix}$$

where lowercase letters indicate _elements_, and the subscript letters indicate the _position of these elements_ in the matrix (and in the table!).

Moreover, any subset of a matrix can be recognized.

We can subset _a row matrix_, as below:

$$\begin{bmatrix}
y_{1,1} & y_{1,2} & \cdots & y_{1,n} \\
\end{bmatrix}$$

We can also subset _a column matrix_, as below:

$$\begin{bmatrix}
y_{1,1} \\ y_{2,2} \\ \vdots \\ y_{m,2}
\end{bmatrix}$$

## Association matrices

Two important matrices can be derived from the ecological data matrix: the _**association matrix among objects**_ and the _**association matrix among descriptors**_.

Using the data from our matrix $Y$,

<div class="math">
\[ Y = 
\begin{array}{cc}
\begin{array}{ccc}
x_1 \rightarrow\\
x_2 \rightarrow\\
\vdots \\
x_m \rightarrow\\
\end{array}
&
\begin{bmatrix}
y_{1,1} & y_{1,2} & \cdots & y_{1,n} \\
y_{2,1} & y_{2,2} & \cdots & y_{2,n} \\
\vdots  & \vdots  & \ddots & \vdots  \\
y_{m,1} & y_{m,2} & \cdots & y_{m,n} 
\end{bmatrix}
\end{array}
\]
</div>

one can examine the relationship between the first two objects:

<div class="math">
\[x_1 \rightarrow \begin{bmatrix}
y_{1,1} & y_{1,2} & \cdots & y_{1,n} \\
\end{bmatrix}
\]
</div>

<div class="math">
\[x_2 \rightarrow 
\begin{bmatrix}
y_{2,1} & y_{2,2} & \cdots & y_{2,n} \\
\end{bmatrix}
\]
</div>

<p>and obtain \(a_{1,2}\). </p>


We can populate the association matrix $A_{n,n}$ with the relationships between all objects from $Y$:

<div class="math">
\[A_{n,n} = 
\begin{bmatrix}
a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
\vdots  & \vdots  & \ddots & \vdots  \\
a_{n,1} & a_{n,2} & \cdots & a_{n,n} 
\end{bmatrix}\]
</div>

<p>Because \(A_{n,n}\) has the same number of rows and columns, it is denoted a <i>square matrix</i>.</p> 

<p>Therefore, \(A_{n,n}\) has \(n^2\) elements.</p>

We can also obtain the relationship between the first two descriptors of $Y$, $y_1$ and $y_2$:

$$\begin{bmatrix}
y_{1,2} \\
y_{2,2} \\
\vdots  \\
y_{m,2} 
\end{bmatrix}$$

$$\begin{bmatrix}
y_{1,1} \\
y_{2,1} \\
\vdots  \\
y_{m,1} 
\end{bmatrix}$$

and store it in $a_{1,2}$.

We can populate the association matrix $A_{m,m}$ with the relationships between all descriptors from $Y$:

$$A_{m,m} = 
\begin{bmatrix}
a_{1,1} & a_{1,2} & \cdots & a_{1,m} \\
a_{2,1} & a_{2,2} & \cdots & a_{2,m} \\
\vdots  & \vdots  & \ddots & \vdots  \\
a_{m,1} & a_{m,2} & \cdots & a_{m,m} 
\end{bmatrix}$$

<p>This \(A_{m,m}\) is a <i>square matrix</i>, and it has \(m^2\) elements.</p>

These matrices, \(A_{n,n}\) and \(A_{m,m}\), are the basis of **_Q-mode_** and **_R-mode_** analyses in ecology.

**_R-mode_** constitutes of analyzing the association between descriptors or species, while **_Q-mode_** analyzes the association between OTUs, objects or sites.

# Exploring the dataset

We will use two main data sets in the first part of this workshop.

They come from Verneaux's PhD thesis (1973), where he proposed to use fish species to characterize ecological zones along European rivers and streams.

He collected data at **30 localities** along the Doubs river, which runs near the France-Switzerland border, in the Jura Mountains.

He showed that fish communities were biological indicators of these water bodies.

Their data is split in three matrices:

1. The abundance of 27 fish species across the communities (`DoubsSpe.csv` and hereon, the `spe` object);
2. The environmental variables recorded at each site (`DoubsEnv.csv` and hereon, the `env` object); and,
3. The geographical coordinates of each site.

Verneaux, J. (1973) _Cours d'eau de Franche-Comté (Massif du Jura). Recherches écologiques sur le réseau hydrographique du Doubs_. Essai de biotypologie. Thèse d'état, Besançon. 1–257.

## Doubs river fish communities 

You can download these datasets from [r.qcbs.ca/workshops/r-workshop-09](http://r.qcbs.ca/workshops/r-workshop-09/).

We can load their data from the `data/` directory in this workshop:

```{r}
spe <- read.csv("data/doubsspe.csv", 
                row.names = 1) 

env <- read.csv("data/doubsenv.csv", 
                row.names = 1)
```

Their data can also be retrieved from the `ade4` package:
```{r eval=FALSE, echo = TRUE}
library (ade4)
data (doubs)

spe <- doubs$fish
env <- doubs$env
```

Alternatively, from the `codep` package:
```{r eval = FALSE, echo = TRUE}
library (codep)
data (Doubs)

spe <- Doubs.fish
env <- Doubs.env
```

We can then explore the objects containing our newly loaded data. 

Let us peek into the `spe` data:

```{r, echo = T}
head(spe)[, 1:8]
```

We can also use the `str()` function, which we learned in Workshops 1 and 2:

```{r, echo = T, output.lines=1:8}
str(spe)
```

You can also try some of these!

```{r, echo = TRUE, eval = FALSE}
# Try some of these!

names(spe)   # names of objects
dim(spe)     # dimensions

str(spe)     # structure of objects
summary(spe) # summary statistics

head(spe)    # first 6 rows
``` 

## Doubs river environmental data

```{r, echo = T}
str(env)
```

It contains the following variables:

|Variable |Description|
|:--:|:--|
|das|Distance from the source [km]  |
|alt|Altitude [m a.s.l.]  |
|pen|Slope [per thousand]  |
|deb|Mean min. discharge [m<sup>3</sup>s<sup>-1</sup>]  |
|pH|pH of water  |
|dur|Ca conc. (hardness) [mgL<sup>-1</sup>]  |
|pho|K conc. [mgL<sup>-1</sup>]  |
|nit|N conc. [mgL<sup>-1</sup>]  |
|amn|NH₄⁺ conc. [mgL<sup>-1</sup>]  |
|oxy|Diss. oxygen [mgL<sup>-1</sup>]  |
|dbo|Biol. oxygen demand [mgL<sup>-1</sup>]  |

You can also use `summary()` to obtain summary statistics from the variables in `env`:

```{r, eval = F}
summary(env) # summary statistics
```
